### 2012-12-29 manuscript.tex
quase terminando a segunda revisão

Data compression is traditionally used to reduce storage resouces usage and/or transmission costs\cite{salomon}. Compression techniques can be classified into lossy and lossless. Goos examples of lossy data compression are sound (mp3), image (JPEG) and movie (mpeg) compression. In this paper we are concerned exclusively with lossless compression. In lossless compression, no information is lost, during the compression process, and they are the method of choice to compress numbers and text.

Any kind of compression incurs in some computational cost. such costs often have to be paid twice since the data needs to be decompressed to be used for its original purpose. Sometimes computational costs are irrelevant, but the need to decompress for usage, can signify that savings in storage costs are applicable to storage situation since we still require the uncompressed space allocation for usage.

Lossless compression methods usually try to exploit redundancies present in the data in order to find a shorter form for of describe the same information content, for example a dictionary based compression, can only store the positions in which a given word occurs in a document, thus saving the space required to store all its repeatitions\cite{salomon2}.

Most if not all existing lossless compression methods were develop under the following paradigm: produce $\rightarrow$ compress $\rightarrow$ store $\rightarrow$ uncompress $\rightarrow$ use. However in the brave new world of big data, the need to analyze data immediately after its coming into existence became the norm. And this analysis must take place, efficiently, within the confines of (RAM) memory. Given a sufficiently big flow of data, compression an decompression costs may become prohibitive as it will slow down data transmission. Moreover, the need to decompress before analyzing removes any potential gains in memory usage.

With the growth of data and analytical demands, creative solutions are need to not only efficiently store large data as consume it on demand. This phenomenon is present in many areas of aplication, ranging from business to science\cite{lynch}, and is being called the big data phenomenon.

One data type which is very common across areas of application is numerical data. Most common statistical and machine learning algorithms operate over large matrices, frequently making use of linear algebra methods. So it would be very desirable to have a compression method in which the compressed object retained the mathematical properties as the original, i.e. could be subject to the same mathematical operations. The idea of operating with compressed array is not new\cite{yemliha2007compiler}, but has yet to applied consistently to the field of numerical computations. One application which employs a form compression is the sparse matrix linear algebra algorithms\cite{dodson1991sparse}, in this case there is no encoding of the data, but only the non-zero elements of the matrices are stored and operated upon.

Currently, the technique most commonly used when dealing with large matrices for numerical computations, is memory mapping\cite{van2011numpy,big}. In memory mapping the matrix is allocated in a virtual contiguous address space which overflows from memory into disk, this way larger than memory datastructures can be manipulated as if they were in memory.  This technique has a big performance penalty due to much lower access speeds of disk when compared to RAM.

In this paper we present two methods that allow the lossless compression of (numerical) arrays, which do not require decompression to be manipulated mathematically. The method involves the encoding of the numbers as strings of bits of variable length. The methods resemble the arithmetic coding\cite{bodden2007arithmetic} algorithm, but is cheaper to compute and does not require decompression. We describe the process of compression and decompression, and how to operate with the arrays in compressed format. We discuss the efficiency of the compression as a function of the distribution of the elements of the matrix.

@@

The main goal of our algorithm in the context of this paper was to achieved the best compression ratio possible while maintaining mathematical equivalence with the original data regarding all arithmetic operations. In order to achieve this, we decided we needed to maintain the structure of the matrix, i.e., the ability to acess any element given its row $i$ and column $j$. In order to achieve compression we decided to exploit inefficiencies in the conventional way matrices are allocated in memory.

The compression method is as follows. Consider a matrix $M_{r \times c}$ in which $r$ is the rows number and $c$ the columns number. Each element of this matrix, called $m_{ij}$, is an integer. Since all information is stored as binary code in digital computers, the conventional way to store arrays of integers in most programming languages is to chose a maximum size for the elements (up to 64bits in a typical CPU), and then allocating a block of memory sufficient to store all the elements of $M$:

\begin{equation}\label{eq:01}
  \mathcal{B} = r \times c \times 64
\end{equation}

The number of bits allocated $\mathcal{B}$, is larger than the absolute minimum number of bits required to represent all the elements of $M$. To help illustrate this fact let's consider as an extreme example: a matrix composed exclusively of 0s and 1s. If we set the element to be 64-bit integers we willbe wasting 63 bits per element of the matrix, since the number of bits really needed to store the this matrix, $\mathfrak{b}$ is given by:

\begin{equation}\label{eq:02}
  \mathfrak{b} = l \times c \times 1
\end{equation}

The potential economy of bits $\xi$ can be represented by the difference between $\mathcal{B}$ and $\mathfrak{b}$ (Equations \ref{eq:03} to \ref{eq:06}).

\begin{equation}\label{eq:03}
 \xi = \mathcal{B} - \mathfrak{b}
\end{equation}

\begin{equation} \label{eq:06}
 \xi =  r \times c \times 63
\end{equation}

From equations \ref{eq:01} to \ref{eq:06}, it can be seen that for any matrix whose greatest element requires less than 64 bits (or the fixed type of the matrix) to be represented, memory savings $\xi$ grow linearly with the size of the matrix.

Now we will introduce two simple methods to exploit these inefficiencies in conventional memory allocation, and save a lot of memory in the process. In this section we will use capital roman letters to denote the original uncompressed matrices and the corresponding lower case letter for the compressed version.

\subsubsection*{Method 1: The Supreme Minimum}

The first methodology consists in determining the value of the greatest element of matrix $M$, which coincides with its supremum, $max M == \sup M$  and determine the minimum number of bits required to store it ($b$) .

\begin{equation} \label{eq:07}
 b(\sup M) \approx \begin{cases}
	1, &  \text{ if } \sup M \in \{0,1\} \\ 
	\lfloor \log_2(\sup M)  \rfloor + 1,  & \text{ if } \sup M > 0  
	\end{cases}
\end{equation}

The allocation of memory still happens in the usual way, only that now, in the space required for a single 64 bit integer, we can store an entire $8\times 8$ matrix of $0$s and $1$s.

To make it clearer, lets look at a concrete example: suppose that the greatest value to be stored in a matrix $M$ is $\max M=1023$. Therefore, the number of bits required to represent it is $10$ $(1111111111)$. Let the first 8 elements of $M$ be:

@@

These elements of $M$, in binary, are shown in Table \ref{tab:01}. It is evident that the number of bits required to represent any other element must be $\leq 10$.

@@

Now consider a matrix $m$, which corresponds to $M$ converted to base $2$ ($M_2$). How do we store such matrix in memory? The conventional way to allocate matrices in memory is to unravel its structure by column (column major, e.g. in Fortran) or by row (row major, e.g. in C) and line them up in a contiguous block of memory. Each element of the matrix is allocated in turn according to the type associated with the matrix.

For our example, since we have determined that each element will require at most 10 bits, we can divide the memory block corresponding to a single 64 bit integer into 6 10-bit chunks which can each hold a single element of $M$. Each of these blocks we will call a bitstring. The remaining 4 bits will be used later. The number of bit strings needed will be $\lfloor \frac{\dim(M) *b} {64} \rfloor +1$.

The final layout of the first 6 elements of $m$ in the first bitstring can be seen in equation \ref{eq:09}.

\begin{equation}\label{eq:09}
 bitstring 1 = 0000\underbrace{0000001010}_{10}\underbrace{0000000001}_{1}\underbrace{0100000000}_{256}\underbrace{1011010001}_{721}\underbrace{1111111111}_{1023}\underbrace{1110000100}_{900} 
\end{equation}

Here is a step-by-step description of the allocation process:
\begin{enumerate}
 \item Element $M_{1,1}=900 = m_{1,1} = 1110000100$ is stored in the first 10-bit chunk of the element strip $bitstring[1]$ , which corresponds to bits 0 to 9.
 \item Element $M_{1,2}=1023$ is allocated in the second chunk, from bit 10 to bit 19.
 \item And so forth: Elements $M_{1,i}$ with $i=1,\ldots,6$ re stored on the remaining chunks (eq. \ref{eq:10}).
 \item Element 7 does not fit on the remaining 4 bits of the first bitstring. So it will straddle two bitstrings, i.e., we will start writing it on the first bitstring and finish on the second (equation \ref{eq:11}).
\end{enumerate}

@@

\begin{equation}\label{eq:10}
  \underbrace{\underbrace{0000010100}_{20}\overbrace{101011}^{a}}_{\phi_{1,2}}|\underbrace{\overbrace{1100}^{b}\underbrace{0000001010}_{10}\underbrace{0000000001}_{1}\hdots}_{\phi_{1,1}}
\end{equation}

\begin{equation}\label{eq:11}
  \underbrace{0000010100}_{20}\underbrace{\overbrace{101011}^a|\overbrace{1100}^{b}}_{700}\underbrace{0000001010}_{10}\ldots
\end{equation}

In conclusion, the final space required to store matrix $m = M_2$ is much smaller than would be required for the conventional storage of $M$ as a 64-bit integer array. In the results section, we will examine in detail the efficiency of this method on real use cases.

\subsubsection*{Method: Variable Bit Length}

In method 1, there is still waste of space since for elements smaller than the supremum, a number of bits may remain unused.

In this second method, we strive to use absolute minimal number of bits necessary to store each value. However, if we are going to divide the biststrings into variable length chunks, we will also need to reserve some extra bits to represent the size of the chunks, otherwise we cannot recover the elements once they are stored.

To illustrate this method, lets consider again the matrix described in equation \ref{eq:08}. Suppose the largest value stored in it is number 1023. Now instead of assigning one chunk of the bitstring to each element of $m$, we will assign two chunks: the first will store the number of bits required to store the value and the second  will store the respective value. The first region will have a fixed size, in the example cited, 4 bits. These 4 bits are the required space to store the bit-length of $\sup M $, in this case, 10.

To illustrate this second method, lets go through the compression process step-by-step. The largest element of $M$ is 1023. The number of bits required to represent it is 10 which in turn requires 4 bits to be stored. Thus the fixed size chunk is 4 bits long for every element.

\begin{enumerate}
 \item The first element $M_{1,1}=900$ requires 10 bits to store, so we write $10_2$ in the first chunk and $900_2$ in the second (Eq. \ref{eq:13}).
 \item We apply the same logic with the following element, $M_{1,2}=1023$.
 \item Element $M_{1,2}=721$ is also added taking the bitstring to the state illustrated in \ref{eq:14}.
\end{enumerate}

\begin{equation} \label{eq:13}
 bitstring_1 = 00000000000000000000000000000000000000000000000000\underbrace{\underbrace{1110000100}_{\text{value} = 900}\underbrace{1010}_{\text{bitlength=10}}}_{900}
\end{equation}

\begin{equation} \label{eq:14}
 bitstring_1 = 
 0000000000000000000000\underbrace{1011010001}_{721}\underbrace{1010}_{10}\underbrace{1111111111}_{1023}\underbrace{1010}_{10}\underbrace{1110000100}_{900}\underbrace{1010}_{10}
\end{equation}

So far the second method is proving more wasteful than the first, but when we add $M_{1,4} =256$  we start to save some space.

\begin{enumerate}
 \item[4.] Element $M_{1,4} =256$ is added.  
 \item [5.] Elements $M_{1,5} =1$ and $M_{1,6} =1$ are added requiring a total of 13 bits instead of 20 with the first method (eq. \ref{eq:16} and \ref{eq:17} ).
 \item [6.] The remaining two elements are added $M_{1,7} =1$ and $M_{1,8} =1$ (eq. \ref{eq:18}).
\end{enumerate}

\begin{equation} \label{eq:16}
 bitstring_1 =
 \underbrace{0100}_{4}\underbrace{1}_{1}\underbrace{0001}_{1}\underbrace{100000000}_{256}\underbrace{1001}_{9}\underbrace{1011010001}_{721}\underbrace{1010}_{10}\underbrace{1111111111}_{1023}\underbrace{1010}_{10}\underbrace{1110000100}_{900}\underbrace{1010}_{10}
\end{equation}

\begin{equation} \label{eq:17}
 bitstring_2 = 000000000000000000000000000000000000000000000000000000000000\underbrace{1010}_{10}
\end{equation}

\begin{equation} \label{eq:18}
 bitstring_2 = 00000000000000000000000000000000000\underbrace{10100}_{20}\underbrace{0101}_{5}\underbrace{1010111100}_{700}\underbrace{1010}_{10}\underbrace{1010}_{10}
\end{equation}

 We used a total of 87 bits to store matrix $m$ with the second method instead of 80 bits using method 1. However, method 2 will be the most efficient for most matrices as show below.

 \subsection*{Efficiency of Compression}
 To compare the two compression methods proposed, we can calculate their efficiencies. However since they depend of the data being compressed, we will describe below how to  calculate the efficienty given the data for both methods. Figure

 \subsubsection*{Method 1}
 Let $M_{r \times c}$ be the matrix we wish to compress. we can define the efficiency of the compression as:
 \begin{equation}\label{eq:eff1}
  \eta_1=\frac{bits\, alocated-bits\, used}{bits\,alocated}
 \end{equation}

@@

% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}
We would like to thank Paulo Cezar Carvalho and Moacyr silva for fruitful discussions and key ideas which helped to improve this paper.

>>>>

Data compression is traditionally used to reduce storage resources usage and/or transmission costs\cite{salomon}. Compression techniques can be classified into lossy and lossless. Examples of lossy data compression are MP3 (audio), JPEG (image) and MPEG (video). In this paper we discuss the use of lossless compression for numerical data structures such as numerical arrays to achieve compression without losing the mathematical properties of the original data.

Lossless compression methods usually exploit redundancies present in the data in order to find a shorter form of describing the same information content. For example, a dictionary-based compression, only stores the positions in which a given word occurs in a document, thus saving the space required to store all its repetitions\cite{salomon2}.

Any kind of compression incurs some computational cost. Such costs often have to be paid twice since the data needs to be decompressed to be used for its original purpose. Sometimes computational costs are irrelevant, but the need to decompress for usage, can signify that the space saved with compression must be available when data is decompressed for usage, thus partially negating the advantages of compression.

Most if not all existing lossless compression methods were developed under the following usage paradigm: \textit{produce} $\rightarrow$ \textit{compress} $\rightarrow$ \textit{store} $\rightarrow$ \textit{uncompress} $\rightarrow$ \textit{use}. The focus of the present work is to allow a slightly different usage:  \textit{produce} $\rightarrow$ \textit{compress} $\rightarrow$ \textit{perform mathematical manipulations} $\rightarrow$ \textit{decompress} (only for human reading).

With the growth of data volumes and analytical demands, creative solutions are needed to efficiently store as well as consume it on demand. This issue is present in many areas of application, ranging from business to science\cite{lynch}, and is being called the \textit{Big Data} phenomenon. In the world of Big Data, the need to analyze data immediately after its coming into existence became the norm. And this analysis must take place, efficiently, within the confines of (RAM) memory. This kind of analyses are what is now known as streaming data analysis\cite{gaber2005mining}. Given a sufficiently dense stream of data, compression an decompression costs may become prohibitive. So having a way to compress data and keeping it compressed for the entire course of the analytical pipeline, is very desirable.

This paper will focus solely on numerical data which for the purpose of the applications is organized as   matrices. This is a most common data structure found in computational data analysis environments. The matrices compressed according to the methodologies proposed here should be able to undergo the same mathematical operations as the original uncompressed matrices. e.g. linear algebra manipulations. This way, the cost of compression is reduced to a single event of compression and no need of decompression except when displaying the results for human reading. The idea of operating with compressed arrays is relatively new\cite{yemliha2007compiler}, and has yet to find mainstream applications to the field of numerical computations. One application which employs a form of compression is the sparse matrix linear algebra algorithms\cite{dodson1991sparse}, in this case there is no alteration in the standard encoding of the data, but only the non-zero elements of the matrices are stored and operated upon.

Larger than RAM data structures can render traditional analytical algorithms impracticable. Currently, the technique most commonly used when dealing with large matrices for numerical computations, is memory mapping\cite{van2011numpy,big}. In memory mapping the matrix is allocated in a virtual contiguous address space which extends from memory into disk. Thus, larger than memory data structures can be manipulated as if they were in memory.  This technique has a big performance penalty due to lower access speeds of disk when compared to RAM.

In this paper we present two methods for the lossless compression of (numerical) arrays. The methods involve the encoding of the numbers as strings of bits of variable length. The methods resemble the arithmetic coding\cite{bodden2007arithmetic} algorithm, but is cheaper to compute. We describe the process of compression and decompression, and study their efficiency under different applications. We also discuss the efficiency of the compression as a function of the distribution of the elements of the matrix.

@@

To maintain mathematical equivalence with the original data for any arithmetic operations, we need to  maintain the structure of the matrix, i.e., the ability to acess any element given its row $i$ and column $j$ and also the numeric nature of its elements. In order to achieve compression we decided to exploit inefficiencies in the conventional way matrices are allocated in memory. The exemples in this paper will be restricted to matrices with positive integer elements.

The compression method is as follows. Let $M_{r \times c}$ be a matrix, in which $r$ is the number of rows and $c$ the number of columns. Each element of this matrix, called $m_{ij}$, is a positive integer. In digital computers, all information is stored as binary code (base 2 numbers). However the conventional way to store arrays of integers is on a sequence of memory blocks of fixed size (power of 2 number of bits), one for each element. The maximum size of a block is equal to the word size of the processor, which for most current CPUs is 64 bits. Some special number such as complex number may be encoded as two blocks instead of one. The size of the chunk of memory allocated to each number will determine their maximum  size (for integers) or their precision (for floating-point numbes). So for matrix $M$, the total memory allocated, assuming chunks of 64 bits, is given by $\mathcal{B} = r \times c \times 64$.

The number of bits allocated $\mathcal{B}$, is larger than the absolute minimum number of bits required to represent all the elements of $M$, since smaller integers, when converted to base 2, require less digits. From now on, when the numerical base will be explicitly notated when necessary to avoid confusion between binary and decimal integers.

Let's consider an extreme example: a matrix composed exclusively of 0s and 1s (base 10). If the matrix type is set to 64-bit integers, 63 bits will be wasted per element of the matrix, since the minimum number of bits needed to store such a matrix is $\mathfrak{b} = r \times c \times 1$. The potential economy of bits $\xi$ can be represented by $\xi = \mathcal{B} - \mathfrak{b} =  r \times c \times 63$.

So it is evident that for any matrix whose greatest element requires less than 64 bits (or the fixed type of the matrix) to be represented, potential memory savings will grow linearly with the size of the matrix.

\subsubsection*{Method 1: The Supreme Minimum (SM)}

The SM method consists in determining the value of the greatest element of matrix $M$, which coincides with its supremum, $max M == \sup M$  and determine the minimum number of bits, $b(\sup M$), required to store it. We will use capital roman letters to denote uncompressed matrices and the corresponding lower case letter for the compressed version.

\begin{equation} \label{eq:07}
 b(\sup M) \approx \begin{cases}
	1, &  \text{ if } \sup M \in \{0,1\} \\ 
	\lfloor \log_2(\sup M)  \rfloor + 1,  & \text{ if } \sup M > 1  
	\end{cases}
\end{equation}

The allocation of memory still happens in the usual way, i.e. in fixed size 64-bit chunks, only that now, in the space required for a single 64 bit integer, we can store for example, an entire $8\times 8$ matrix of $0_{10}$ and $1_{10}$.

Let's look at a concrete example: suppose that the greatest value to be stored in a matrix $M$ is $\max M=1023$. Therefore, the number of bits required to represent it is $10$ $(1111111111)$. Let the first 8 elements of $M$ be:

@@

These elements of $M$, in binary, are shown in Table \ref{tab:01}. It is evident that the number of bits required to represent any other element must be $\leq 10$. From now on the minimum number of bits required to represent a base 10 integer will be refered to as its bit-length.

@@

to store matrix $M$ it first has to be converted to base $2$ ($M_2$). Then it will be unraveled by column (column major, e.g. in Fortran) or by row (row major, e.g. in C) and its elements will be written as  fixed size adjacent chunks of memory. The size of each chunk is determined by the type associated with the matrix (typically 64 bits, but always a power of 2).

According to the SM method, having determined that each element will require at most 10 bits, we can divide the memory block corresponding to a single 64 bit integer into 6 10-bit chunks which can each hold a single element of $M$. These 64-bit blocks will be called a bitstring. The remaining 4 bits will be used later. The number of bitstrings needed will be $\lfloor \frac{\dim(M) *b(\sup M)} {64} \rfloor +1$.

The final layout of the first 6 elements of $m$ in the first bitstring can be seen in \ref{eq:09}.

\begin{equation}\label{eq:09}
 bitstring_1 = 0000\underbrace{0000001010}_{10}\underbrace{0000000001}_{1}\underbrace{0100000000}_{256}\underbrace{1011010001}_{721}\underbrace{1111111111}_{1023}\underbrace{1110000100}_{900} 
\end{equation}

Here is a step-by-step description of the aplicationof the SM method to matrix $M$:
\begin{enumerate}
 \item Element $M_{1,1}=900 = m_{1,1} = 1110000100$ is stored in the first 10-bit chunk of the element strip $bitstring[1]$ , which corresponds to bits 0 to 9.
 \item Element $M_{1,2}=1023$ is allocated in the second chunk, from bit 10 to bit 19.
 \item Repeat for elements $M_{1,i}$ with $i=1,\ldots,6$ which are stored on the remaining chunks.
 \begin{equation*}\label{eq:10}
  \underbrace{\underbrace{0000010100}_{20}\overbrace{101011}^{b}}_{bitstring_2}|\underbrace{\overbrace{1100}^{a}\underbrace{0000001010}_{10}\underbrace{0000000001}_{1}\hdots}_{bitstring_1}
\end{equation*}
 \item Element 7 does not fit on the remaining 4 bits of the first bitstring. So it will straddle two bitstrings, i.e., it is divided in two segments $a$ and $b$, $a$ is written on the first bitstring $b$ on the second.
 \begin{equation*}\label{eq:10}
  \underbrace{\underbrace{0000010100}_{20}\overbrace{101011}^{b}}_{bitstring_2}|\underbrace{\overbrace{1100}^{a}\underbrace{0000001010}_{10}\underbrace{0000000001}_{1}\hdots}_{bitstring_1}
\end{equation*}
\end{enumerate}

@@

\begin{equation*}\label{eq:11}
  \underbrace{0000010100}_{20}\underbrace{\overbrace{101011}^a|\overbrace{1100}^{b}}_{700}\underbrace{0000001010}_{10}\ldots
\end{equation*}

Thus the compressed matrix $m = M_2$ requires less memory than the conventional storage of $M$ as a 64-bit integer array.

\subsubsection*{Method 2: Variable Length Blocks (VLB)}

In the SM method, there is still waste of space since for elements smaller than the supremum, a number of bits remain unused.

In the VBL method, the absolute minimal number of bits are used to store each value. However, if we are going to divide the biststrings into variable length chunks, we also need to reserve some extra bits to represent the size of each chunk, otherwise the elements cannot be recovered once they are stored.

Lets use again the matrix described in equation \ref{eq:08}, where  the largest element is number 1023. Now instead of assigning one chunk of the bitstring to each element of $m$, we will assign two chunks: the first will store the number of bits required to store the element and the second  will store the actual element. The first chunk will have a fixed size, in this case, 4 bits. These 4 bits are the required space to store the bit-length of $\sup M $, in this case, 10.

Lets go through VLB compression step-by-step. The largest element of $M$ is 1023. Its bit-length is 10 which in turn is 4 bits long in base 2 (1010). Thus the fixed size chunk is 4 bits long for every element.

\begin{enumerate}
 \item The first element $M_{1,1}=900$ requires 10 bits to store, so we write $10$ in the first chunk and $900$ in the second.
 \begin{equation*} \label{eq:13}
 bitstring_1 = 00000000000000000000000000000000000000000000000000\underbrace{\underbrace{1110000100}_{\text{element} = 900}\underbrace{1010}_{\text{bit-length=10}}}_{M_{1,1}}
\end{equation*}
 \item Do the same for the next element, $M_{1,2}=1023$.
 \item Element $M_{1,2}=721$ is also added taking the bitstring to the state.
 \begin{equation*} \label{eq:14}
 bitstring_1 = 
 0000000000000000000000\underbrace{1011010001}_{721}\underbrace{1010}_{10}\underbrace{1111111111}_{1023}\underbrace{1010}_{10}\underbrace{1110000100}_{900}\underbrace{1010}_{10}
\end{equation*}
\end{enumerate}

So far the VLB method is more wasteful than the SM, but when we add $M_{1,4} =256$  we start to save some space.

\begin{enumerate}
 \item[4.] Element $M_{1,4} =256$ is added.  
 \item [5.] Elements $M_{1,5} =1$ and $M_{1,6} =10$ are added requiring a total of 13 bits instead of 20 with the SM method.
 \begin{align*} \label{eq:16}
 bitstring_1 &=
 \underbrace{0100}_{4}\underbrace{1}_{1}\underbrace{0001}_{1}\underbrace{100000000}_{256}\underbrace{1001}_{9}\underbrace{1011010001}_{721}\underbrace{1010}_{10}\underbrace{1111111111}_{1023}\underbrace{1010}_{10}\underbrace{1110000100}_{900}\underbrace{1010}_{10} \\
 bitstring_2 &= 000000000000000000000000000000000000000000000000000000000000\underbrace{1010}_{10}
\end{align*}
 \item [6.] The remaining two elements are added $M_{1,7} =700$ and $M_{1,8} =20$.
 \begin{equation*} \label{eq:18}
 bitstring_2 = 00000000000000000000000000000000000\underbrace{10100}_{20}\underbrace{0101}_{5}\underbrace{1010111100}_{700}\underbrace{1010}_{10}\underbrace{1010}_{10}
\end{equation*}
\end{enumerate}

 We used a total of 87 bits to store matrix $m$ with the VLB method instead of 80 bits using the SM method. However, as shall be seen later, the VLB method will be the most efficient for most matrices.

 \subsection*{Compression Efficiency}
 Compression efficiency depends of the data being compressed. Below, a formula for calculating compression efficiency is derived for both methods. They will be based on the following ratio:

 \begin{equation}\label{eq:eff1}
  \eta=\frac{bits\, alocated-bits\, used}{bits\,alocated}
 \end{equation}
 \subsubsection*{Method 1}
 Let $M_{r \times c}$ be the matrix we wish to compress. we can define the efficiency of the compression as:

@@

% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}
We would like to thank Claudia Torres Codeço, Paulo Cezar Carvalho and Moacyr silva for fruitful discussions and key ideas which helped to improve this paper.
